---
title: "session07 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: false
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, lme4, brms, bayestestR)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

### Introduction to the problem

We will be working with Quantese again, see `01_exercise.qmd` and the other data resources for a refresher.

```{r}
#| label: load-quantese

quantese <- read_csv("../data/quantese_naming_sim.csv") 
# add log-transformed reaction time
quantese <- quantese |> 
  mutate(log_rt = log(rt))

# have a look
quantese

```

In an earlier homework, we have assessed the following hypothesis:

*"We hypothesize that log-transformed reaction times are affected by lexical frequency"*

We tried to run the following model that included an appropriate random effect structure but it did not converge with `lme4`:

```{r}
#| label: lmer_quantese

# formulate the corresponding LMER
quant_lmer <- lmer(log_rt ~ log_frequency +
                     (1 + log_frequency | PartID) +
                     (1 | ItemID), 
                    data = quantese)
# your receive a warning: "boundary (singular) fit: see help('isSingular')"

```

Now we will re-run this model within the Bayesian framework:

### (A) Specify priors

Check which priors this model needs and specify weakly informative priors.

Remember that weakly informative priors are priors that do not bias the posteriors in any direction, but reasonably constrain the parameter space, i.e. only reasonable values are assumed.

For priors of class `Intercept` and `b`, constrain the parameter space with a wide normal distribution.

```{r}
#| label: get_prior

# definign our model formula
formula <- bf(log_rt ~ log_frequency +
                   (1 + log_frequency | PartID)+
                   (1 | ItemID))
# checking names of parameters
get_prior(formula, quantese, family = gaussian())

# defining my own priors:
# loading package:
library(bayesrules)
# what we know from data set:
# our logfrequency is only positive (between 1 and 6)
# log_rt has to be positive and might not be above 10 (looking at data, it's inbetween 5.8 and 7.8)
# visualizing different normal distributions:
# plot_normal(mean= 5, sd = 2)    # rt of 0 is actually not possible, so maybe, we should assume sth starting from 200ms -> log(200) = 5.298...
plot_normal(mean= 6, sd = 2) # 0 is than still possible

# for the log_frequency/our slope
plot_normal(mean = 0, sd = 3)

# specify weakly informative priors
my_priors <- c(prior(normal(6,2),
                     class = 'Intercept'),
               prior(normal(0,3),
                     class = 'b',
                     coef = 'log_frequency'))

```

We can also now get used to specifying priors for classes `sd` and `cor`. Priors of class `sd` are priors for the variance associated with our random effects. You see here 2 `sd` priors for group `ItemID`. The first one is just a generic prior for all `ItemID` coefficients and the second one with the `coef = Intercept` stands for the random intercept variation between items. Similarly for group `PartID`, we have 3 priors. The first one is again just a generic prior for all `PartID` coefficients. The second one with the `coef = Intercept` stands for the random intercept variation between participants. And the last one with the `coef = log_frequency` stands for the random slope variation between participants for the predictor `log_frequency`. How do we specify these priors. We have usually no idea what to expect from these parameters, so we here also specify very diffuse priors centered on zero. We pick a cauchy distribution (remember? super squashed normal distributions) because they have thick tails. Let's make them super diffuse and pick a `sd = 3` for both items and participants.

```{r}
#| label: add_sd_priors

# add weakly informative priors for sd
my_priors <- c(my_priors, 
               prior(cauchy(0, 3), class = 'sd', group = 'ItemID'),
               prior(cauchy(0, 3), class = 'sd', group = 'PartID'))

```

And finally, if your model as random slopes, they have by default a correlation parameter for intercept and slope. We are not getting into the math here and will follow advice from smart people. We use Lewandowski-Kurowicka-Joe priors (shape = 1): `lkj(1)`. Specifying this prior for random slope models actually helps a lot with the fitting procedure.

```{r}
#| label: add_cor_priors

# add weakly informative priors for cor
my_priors <- c(my_priors, 
               prior(lkj(1), class = 'cor'))

```

### (B) Run model

Now run the model with 4 chains, 1000 warm-up and 1000 post-warm-up samples per chain. Check that chains are reasonably mixed and that sampling yielded a sufficient effective sample size. If not address the issues appropriately (check slides again).

```{r}
#| label: model_quant

# if the model takes way too long to sample (more than 5-10 mins), use the following subset
# quantese_sub <- quantese |> 
#   filter(PartID %in% c("01","02","03","04",
#                        "05","06","07","08"))

# # fit brms model to check names of parameters
quant_brm <- brm(log_rt ~ log_frequency +
                   (1 + log_frequency | PartID)+
                   (1 | ItemID),
                 # specify error distribution
                 family = gaussian(),
                 # adding our priors
                 prior = my_priors,
                 # specify data
                 data = quantese)
# model output
quant_brm

# checking convergence diagnostics from output above
# Assessment:
# For the regression coefficiants Intercept and log_frequency, the Rhat value is 1.01, indicating a potential lack of convergence. In addition, the effective sample sizes both for bulk for each of these two parameters are below 400, indicating that samples might be autocorrelated.
# For the random intercept for participants, the Rhat value is 1.02 and bulk ESS is below 400 too.

# checking chain convergence
mcmc_plot(quant_brm, type = 'trace')
# Assessment
# the caterpillar plots look all good

# checking again for the ones with low ESS values
# loading package
library(bayesplot, tidybayes)
# getting correct name of variables
tidybayes::get_variables(quant_brm)
# lookign at chain convergence for the parameters that had a high Rhat value and low bulk ESS
mcmc_dens_overlay(quant_brm, pars = 'sd_PartID__Intercept')
mcmc_dens_overlay(quant_brm, pars = 'b_Intercept')
mcmc_dens_overlay(quant_brm, pars = 'b_log_frequency')

# non of their plots show a very bad convergence, even though in all three plots, the gap between the different lines increases the closer to the peak the lines are.

pp_check(quant_brm)

# Solutions:
# increase interations and warmup by adding
# iter = 6000,warmup = 4000
# to our brm model 

```

### (C) Plot posterior predictions

Find a way to extract posterior samples from the model for the `Intercept` and the `log_frequency` coefficient and plot the predicted posterior mean linear model alongside its 95% Credible Intervals. There are many different ways to get there.

```{r}
#| label: plot_predictions

#loading package
library(tidybayes)

# get correct names of variables
get_variables(quant_brm)

#plotting posterior for b_Intercept and b_log_frequency

# first attempt - does not include 95% credible interval
quant_brm |>
  spread_draws(b_Intercept,
               b_log_frequency) |>
  ggplot(aes(x = b_log_frequency,
             y = b_Intercept))+
  geom_bin2d()+
  theme_minimal()

# second attempt
quant_brm |>
  spread_draws(b_Intercept,
               b_log_frequency) |>
  ggplot(aes(x = b_log_frequency,
             y = b_Intercept))+
  stat_halfeye()


```
