---
title: "session08 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayesrules, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Exercise

We will be re-analyzing the data from @winter2013making.

```{r}
#| label: preprocessing_sim

# Load our datasets
sim <- read_csv("../data/linguistic_similarity.csv") |> 
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
sim

```

We will now retrace the original article's complete analysis. In their original analysis, they were interested in whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities.

## (a) Specify Priors

Specify the model formula and appropriate weakly informative priors for all relevant predictors. Contrast-code the predictors.

```{r}
#| label: model_formula_sim

# contrastcoding the predictors
sim <- sim |>
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
# double checking levels for both predictors
levels(sim$Sem)
levels(sim$Phon)
# setting new contrasts so that 
contrasts(sim$Sem) <- c(1,0) #Similar = 0, Different = 1
contrasts(sim$Phon) <- c(1,0) #Similar = 0, Different = 1, 

# specifying model formula
mdl_formula <- bf(Distance ~ Sem * Phon)

# checking which priors I need to set and what the coefficients names are
get_prior(mdl_formula, sim, family = "gaussian")
# Visualizing priors to make sure they look like I imagine
plot_normal(108, 35)
# setting my priors
my_priors <- c(prior(normal(108,35), class = "Intercept"),
               prior(normal(0,35), class = "b"))

```

## (b) Run the appropriate model for the research question

```{r}
#| label: model_sim

# fitting and runnin gthe model
mdl_sim <- brm(mdl_formula,
               family = gaussian(),
               prior = my_priors,
               data = sim)

```

## (c) Posterior predicive check

Check the overall model fit with posterior predictive checks. If the model does not fit the data well, fix the problem (check `06_homework_solutions.qmd` for inspiration).

```{r}
#| label: model_fit

# assessing the models fit
# checking Rhat and ESS
mdl_sim
# assessment: all Rhat values are exactly 1.0 and Bulk_ESS above 400 and 
# Tail_ESS above 100

# chould the model have generated the data?
pp_check(mdl_sim, ndraws = 100)
# the data is actually not really normally distributed, looking at the right 
# side of the density curve which is quite wiggly 

# solution: square rooting the response 

sim_mod <- sim |>
  mutate(sqrt_Distance = sqrt(Distance))

# changing response variable name in formula
mdl_formula_mod <- bf(sqrt_Distance ~ Sem * Phon)

# Visualizing priors to make sure they look like I imagine
plot_normal(10.39, 5.92)
# adjusting my priors by square-rooting my initial "108, 35"
my_priors_mod <- c(prior(normal(10.39, 5.92), class = "Intercept"),
               prior(normal(0, 5.92), class = "b"),
               # Also forgot to specify sigma for the original model
               # but am doing it now for the adjusted one
               prior(normal(0, 3), class = "sigma"))

# rerunning the model
mdl_sim_mod <- brm(mdl_formula_mod,
               family = gaussian(),
               prior = my_priors_mod,
               data = sim_mod)

# assessing new model
mdl_sim_mod #comment: Rhat and ESS values just as good as before :)
pp_check(mdl_sim_mod, ndraws= 100)
# now the black and light blue ines match a lot better and I will proceed with the adjusted model mdl_sim_mod

```

## (d) Assess sampling diagnostics

Check that chains are reasonably mixed and that sampling yielded a sufficient number of effective sample size. If there are issues, address them accordingly.

```{r}
#| label: sampling_diagnostics

# through caterpillarplots
mcmc_plot(mdl_sim_mod,
          type = "trace")
# Looks like 5 healthy caterpilars


# checking correct names for coefficients
get_prior(mdl_formula_mod, sim_mod, family = "gaussian")
# checking how the chains overlap for all four parameters
mcmc_dens_overlay(mdl_sim_mod,
                  pars = 'Intercept')
mcmc_dens_overlay(mdl_sim_mod,
                  pars = 'b_Phon1')
mcmc_dens_overlay(mdl_sim_mod,
                  pars = 'b_Sem1')
mcmc_dens_overlay(mdl_sim_mod,
                  pars = 'b_Sem1:Phon1')
mcmc_dens_overlay(mdl_sim_mod,
                  pars = 'sigma')

# All 5 look fine
```

## (e) Interpret the results

Now interpret the model relative to the research question and describe uncertainty, visualize posteriors, calculate posterior probabilities of direction, and calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.

(Note: When you want to extract interaction terms like `b_Sem1:Phon1`, you have to put single quotation marks around them because R struggles with interpreting the colon)

```{r}
#| label: interpret_sim

# model interpretation
mdl_sim_mod

# The model predicts an average difference between semantically and phonetically similar cities of 8.29 [7.79,8.79], which is 69mm with a 95% credibility interval ranging from 61mm to 77mm.
# For phonetically similar and semantically different cities, distance increases by 0.78 [0.10,1.47], with an average distance of 82mm.
# For semantically similar and phonetically different cities, however, distance decreases by -0.34 [-1.06,0.34], with an average distance of 63mm, thus being closer to each other than phonetically similar cities.
# If cities are both phonetically and semantically different, the average distance is 73mm due to the interaction term of -0.19 [-1.18,0.83]. Also, the interaction term shows a standard error twice of the sie of the one for the other two predictors.


# visualizing the posterior distribution
# posterior for Semantic Properties
post_Sem <- mdl_sim_mod|>
  spread_draws(b_Sem1)|>
  ggplot(aes(x = b_Sem1)) +
  stat_halfeye(fill = "orange") +
  geom_vline(xintercept = 0,
             lty = 'dashed') +
  scale_x_continuous()+
  labs(title = "\nPosterior distribution of phonologically similar and\nsemantically different cities",
       x = "\nDistance in sqrt(mm)",
       y = "") +
  scale_x_continuous(limits = c(-2,2))+
  theme_minimal()
post_Sem

# posterior for Phonetical Properties
post_Phon <- mdl_sim_mod|>
  spread_draws(b_Phon1)|>
  ggplot(aes(x = b_Phon1)) +
  stat_halfeye(fill = "skyblue") +
  geom_vline(xintercept = 0,
             lty = 'dashed') +
  scale_x_continuous()+
  labs(title = "\nPosterior distribution of semantically similar and\nphonologically different cities",
       x = "\nDistance in sqrt(mm)",
       y = "") +
  scale_x_continuous(limits = c(-2,2))+
  theme_minimal()
post_Phon

# posterior for Interaction Term
post_Int <- mdl_sim_mod|>
  spread_draws(`b_Sem1:Phon1`)|>
  ggplot(aes(x = `b_Sem1:Phon1`)) +
  stat_halfeye(fill = "purple") +
  geom_vline(xintercept = 0,
             lty = 'dashed') +
  scale_x_continuous()+
  labs(title = "\nPosterior distribution of phonologically and\nsemantically different cities",
       x = "\nDistance in sqrt(mm)",
       y = "") +
  scale_x_continuous(limits = c(-2,2))+
  theme_minimal()
post_Int


# posterior probabilities of direction
mdl_sim_mod|>
    spread_draws(b_Sem1)|>
    select(b_Sem1)|>
    p_direction()
# p(b<0) = 0.9883


mdl_sim_mod |>
    spread_draws(b_Phon1) |>
    select(b_Phon1) |>
    p_direction()
# p(b<0) = 0.832

mdl_sim_mod |>
    spread_draws(`b_Sem1:Phon1`) |>
    select(`b_Sem1:Phon1`) |>
    p_direction()
# p(b<0) = 0.6435


#  Bayes Factors
# fitting model without data
mdl_sim_mod_priors <- brm(mdl_formula_mod,
               family = gaussian(),
               prior = my_priors_mod,
               sample_prior = "only",
               data = sim_mod)

# for Sem
# extracting posteriors of the models coefficients
posteriors_Sem <- mdl_sim_mod |>
  spread_draws(b_Sem1) |>
    select(b_Sem1)
# extracting priors of model
priors_Sem <- mdl_sim_mod_priors |>
  spread_draws(b_Sem1) |>
    select(b_Sem1)
# running bayes factor against null value
bayesfactor_parameters(posteriors_Sem,
                       priors_Sem,
                       direction = "two-sided",
                       null = 0)
# bf = 0.673

# for Phon
# extracting posteriors of the models coefficients
posteriors_Phon <- mdl_sim_mod |>
  spread_draws(b_Phon1) |>
    select(b_Phon1)
# extracting priors of model
priors_Phon <- mdl_sim_mod_priors |>
  spread_draws(b_Phon1) |>
    select(b_Phon1)
# running bayes factor against null value
bayesfactor_parameters(posteriors_Phon,
                       priors_Phon,
                       direction = "two-sided",
                       null = 0)
# bf = 0.105

# for Interaction Term
# extracting posteriors of the models coefficients
posteriors_Int <- mdl_sim_mod |>
  spread_draws(`b_Sem1:Phon1`) |>
    select(`b_Sem1:Phon1`)
# extracting priors of model
priors_Int <- mdl_sim_mod_priors |>
  spread_draws(`b_Sem1:Phon1`) |>
    select(`b_Sem1:Phon1`)
# running bayes factor against null value
bayesfactor_parameters(posteriors_Int,
                       priors_Int,
                       direction = "two-sided",
                       null = 0)
# bf = 0.086

```

## (g) Write up your entire analysis:

I forgot to set a seed, so the actual results when running my code may differ slightly from what I report here and throughout my script:

The model posterior for semantically and phonologically similar cities, the intercept, is 69mm. The square-root transformed model output is 8.29 with a 95% Credible Interval of [7.79,8.79], ranging from 61mm to 77mm.

The models posterior distributions indicate, that sqaure-root transformed distance in between cities increase on average by 0.78 when two cities are semantically different (and phonologically similar) with a 95% credible interval of [0.10,1.47]. The average distance in mm is 82.
Given our model, our data and our priors, 98.83% of the posterior sample is positive, showing that negative values ae possible, but not very likely.

For phonetically different (and semantically similar) cities, the sqaure-root transformed distance decreased on average by -0.34 [95% CI: -1.06,0.34]. The average distance in mm is 63.
Given our model, our data and our priors, 80.30% of the posterior sample is negative, which indicates that both negative and positive values are possible, even though negative ones are more likely.

The interaction term, square-root transformed, is -0.19 with a 95% credible interval ranging [-1.18,0.83]. The average distance of both semantically and phonologically different cities is 73mm.  The interaction term shows a standard error twice of the size of the one for the other two predictors (0.49).
Given our model, our data and our priors, 64.35% of the posterior sample is negative, which indicates that both negative and positive values are possible, with negative values being slightly more likely compared to positive ones.

Subsequently, we calculate the Savage-Dickey density ratio [@DickeyLientz1970-The-Weighted-Li; @WagenmakersLodewyckx2010-Bayesian-hypoth; @roettgermeaningful], i.e., we relate the evidence (the probability of the posterior distribution at point-zero) for the model based on the priors only (i.e., before seeing the data) to the evidence for the model based on both priors and likelihood (i.e., after seeing the data) [e.g., @WetzelsGrasman2010-An-encompassing]. The Savage-Dickey method lets us assess evidence for and against a null hypothesis using Bayes factor. Here, the Bayes factor indicates the degree by which the mass of the posterior distribution has shifted further away from the point null value (relative to the prior distribution).
(Thank you Timo for the nice wording).

For Sem, the analysis yields a Bayes Factor of 0.673, indicating that the probability of the effect of semantical difference being point null has increased a tiny little bit, suggesting anecdotal evidence against the alternative hypothesis that the effect of semantical difference is point null[@lee-wagenmakers_Bayesian_book_2014].

For Phon, the analysis yields a Bayes Factor of 0.105, indicating that the probability of the effect of ponological difference being point null has increased again only slighly, suggesting moderate evidence against the alternative hypothesis that the effect of phonological difference is point null[@lee-wagenmakers_Bayesian_book_2014].

For the interaction term, the analysis yields a Bayes Factor of 0.086, indicating that the probability of the effect of the interaction of semantical and phonological difference being point null has increased, suggesting strong evidence against the alternative hypothesis that the effect of the interaction of semantical and phonological difference difference is point null[@lee-wagenmakers_Bayesian_book_2014].

However, when calculating the Bayes Factors, the suggested sample size of at least 40.000 samples was by far not reached. Only 4000 samples where used.


## OPTIONAL (h) compare your interpretation to the original findings

YOUR COMPARISON HERE

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
